ML Estimation of Matrix--$t$ Distributions with Expectation Maximization

### 2017-01-21 paper.tex
Start modifying the M-step.

Given a random variable $X$ with probability distribution with density (or mass) function $p_{\theta}(\cdot)$ with $\theta$ dependence. Assume the density (or mass) function is a marginalization over some unobserved random state $Z$, i.e., $p_{\theta}(x)=\mathbb{E}_{\theta}\left[p_{\theta}(x|Z)\right]$. With observations $\left\{x_i\right\}_{i\in I}$ the classical EM method of MLE consists of two steps:

@@

We can split the integral in Eq.~\eqref{eq:expdef} by the terms of $\log f(T_{j},S;\theta)$. Eliminating a global factor $\frac{1}{2}$ and terms which do not depend on the parameters in $\theta$ we have
\begin{equation}
\begin{split}\log f(T_{j},S;\theta)\sim -\nu\log|S|-\textrm{tr}\left[S^{-1}\left\{\Sigma+(T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}}\right\}\right]\\
-\nu p\log2-p\log |\Omega|+(\nu+p-1)\log |\Sigma|-2\log\Gamma_{p}\left(\frac{\nu+p-1}{2}\right).
\end{split}\label{eq:logf}
\end{equation}
The integrals to be computed are
\begin{equation}
\int_{\mathbf{S}_+^p}f(T_j,S;\theta_k)\log |S|\,dS, \qquad 
\int_{\mathbf{S}_+^p}f(T_j,S;\theta_k) S^{-1}\,dS
\end{equation}
and they can be done by using the equalities derived by taking the derivative of Eq.~\eqref{eq:intmultgammadef} with respect to $\alpha$ and matrix $A$, respectively,
\begin{equation}
\int_{\mathbf{S}_+^p}e^{-\textrm{tr}\left[SA\right]}|S|^{\alpha}\log |S|\,dS=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}}\left\{\gamma_{p}\left(\alpha+\frac{1+p}{2}\right)-\log |A|\right\},\label{eq:intderivative1}
\end{equation}
\begin{equation}
\int_{\mathbf{S}_+^p}e^{-\textrm{tr}\left[SA\right]}|S|^{\alpha}S\,dS=\left(\alpha+\frac{1+p}{2}\right)\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}}A^{-1}.\label{eq:intderivative2}
\end{equation}

Utilizing the equalities \eqref{eq:intderivative1} and \eqref{eq:intderivative2} with adequate rearrangements we obtain the closed form expression of $Q_j(\theta,\theta_k)$ while ignoring the irrelevant constant terms
\begin{equation}
\begin{split}Q_j(\theta,\theta_k) \sim \nu\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)-2\log\Gamma_{p}\left(\frac{\nu+p-1}{2}\right)-p\log|\Omega|\\
-\nu\log|\Sigma_k+(T_j-M_k)\Omega_k^{-1}(T_j-M_k)^{\textrm{t}}|+(\nu+p-1)\log |\Sigma|\\
-(m+\nu_k+p-1)\textrm{tr}\left[\left\{\Sigma_k+(T_j-M_k)\Omega_k^{-1}(T_j-M_k)^{\textrm{t}}\right\}^{-1}\right.\\
\left.\left\{(T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}}+\Sigma\right\}\right].
\end{split}
\label{eq:fsimp}
\end{equation}
The maximization formulae follows by setting the gradient of the parameters to zero.

For $M_{t}$ we have:
\begin{equation}
\frac{dE}{dM_{t}}=0\Rightarrow\sum_{i}^{N}\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}\left(T_{i}-M_{t}\right)\Omega_{t}^{-1}=0
\end{equation}
and because $\Omega_{t}$ is nonsingular, we can obtain $M_{t}$ as
a weighted average of the input data:
\begin{equation}
M_{t}=\frac{1}{N}\sum_{i}^{N}\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}T_{i}=\frac{1}{N}\sum_{i}^{N}W_{i}T_{i}
\end{equation}
where we have defined the weight matrices $W_{i}$ as
\begin{equation}
W_{i}=\left((T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}+\Sigma\right)^{-1}
\end{equation}

$\Sigma_{t}$ is given in terms of the inverse of the mean weight
by:
\begin{equation}
\Sigma_{t}=\frac{n_{t}+p-1}{n+m+p-1}\left[\frac{1}{N}\sum_{i}^{N}W_{i}\right]^{-1}
\end{equation}
Note that it depends on $n_{t}$.

For $\Omega_{t}$ we obtain:
\begin{equation}
\Omega_{t}=\frac{n+p+m-1}{p}\frac{1}{N}\sum_{i}^{N}(T_{i}-M_{t})W_{i}(T_{i}-M_{t})
\end{equation}
and for $n_{t}$, in the case where $\Sigma_{t}$ is a free parameter
that needs to be estimated, we have:
\begin{equation}
-p\log(m+n+p-1)+p\log(n_{t}+p-1)+\frac{1}{N}\sum_{i}^{N}\log|W_{i}|-\log\det\left|\frac{1}{N}\sum_{i}^{N}W_{i}\right|+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
\end{equation}
otherwise if $\Sigma_{t}=\Sigma$ is fixed and does not depend on
$n_{t}$:
\begin{equation}
\log\det\Sigma_{t}-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
\end{equation}

>>>>

Given a random variable $X$ with probability distribution with density (or mass) function $p_{\theta}(\cdot)$ with $\theta$ dependence. Assume the density (or mass) function is a marginalization over some unobserved random state $Z$, i.e., $p_{\theta}(x)=\mathbb{E}_{\theta}\left[p_{\theta}(x|Z)\right]$. With observations $\left\{x_i\right\}_{i\in I}$ the classical EM method for MLE consists of two steps:

@@

We can split the integral in Eq.~\eqref{eq:expdef} by the terms of $\log f(T_{j},S;\theta)$. Eliminating a global factor $\frac{1}{2}$ and terms which do not depend on the parameters in $\theta$ we have
\begin{equation}
\begin{split}\log f(T_{j},S;\theta)\sim -\nu\log|S|-\textrm{tr}\left[S^{-1}\left\{\Sigma+(T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}}\right\}\right]\\
-\nu p\log2-p\log |\Omega|+(\nu+p-1)\log |\Sigma|-2\log\Gamma_{p}\left(\frac{\nu+p-1}{2}\right).
\end{split}\label{eq:logf}
\end{equation}
The integrals to be computed are
\begin{equation}
\int_{\mathbf{S}_+^p}f(T_j,S;\theta_k)\log |S|\,dS, \qquad 
\int_{\mathbf{S}_+^p}f(T_j,S;\theta_k) S^{-1}\,dS.
\end{equation}
They can be done by using the equalities derived by taking the derivative of Eq.~\eqref{eq:intmultgammadef} with respect to $\alpha$ and matrix $A$, respectively,
\begin{equation}
\int_{\mathbf{S}_+^p}e^{-\textrm{tr}\left[SA\right]}|S|^{\alpha}\log |S|\,dS=\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}}\left\{\gamma_{p}\left(\alpha+\frac{1+p}{2}\right)-\log |A|\right\},\label{eq:intderivative1}
\end{equation}
\begin{equation}
\int_{\mathbf{S}_+^p}e^{-\textrm{tr}\left[SA\right]}|S|^{\alpha}S\,dS=\left(\alpha+\frac{1+p}{2}\right)\Gamma_{p}\left(\alpha+\frac{1+p}{2}\right)|A|^{-\alpha-\frac{1+p}{2}}A^{-1}.\label{eq:intderivative2}
\end{equation}

Utilizing the equalities \eqref{eq:intderivative1} and \eqref{eq:intderivative2} with adequate rearrangements we obtain the closed form expression of $Q_j(\theta,\theta_k)$ while ignoring the irrelevant constant terms
\begin{equation}
\begin{split}Q_j(\theta,\theta_k) \sim \nu\left\{\log|W_j^{(k)}|+\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)\right\}-2\log\Gamma_{p}\left(\frac{\nu+p-1}{2}\right)-p\log|\Omega|\\
-(m+\nu_k+p-1)\textrm{tr}\left[W_j^{(k)}\left\{\Sigma+(T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}}\right\}\right]+(\nu+p-1)\log |\Sigma|,
\end{split}
\label{eq:fsimp}
\end{equation}
where the matrices $W_j^{(k)}$ are defined as
\begin{equation}
W_j^{(k)}=\left\{\Sigma_k+(T_j-M_k)\Omega_k^{-1}(T_j-M_k)^{\textrm{t}}\right\}^{-1}.
\end{equation}

To maximize $Q(\theta,\theta_k)$ with respect to $\theta$ we compute the gradient of each parameter and equate them to $0$ to obtain $\theta_{k+1}$. For parameter $M$ we have
\begin{equation}
\sum_{j\in J}W_j^{(k)}\left(T_j-M\right)\Omega^{-1}=0.
\end{equation}
As $\Omega$ is nonsingular we obtain $M_{k+1}$ as an average of the input data weighted by $W_j^{(k)}$
\begin{equation}
M_{k+1}=\left(\sum_{j\in J}W_j^{(k)}\right)^{-1}\sum_{j\in J}W_j^{(k)}T_j.
\end{equation}
Maximizing $Q(\theta,\theta_k)$ with respect to $\Sigma$ we obtain $\Sigma_{k+1}$ in terms of the inverse of the mean of $W_j^{{k}}$
\begin{equation}
\Sigma_{k+1}=\frac{\nu+p-1}{\nu_k+m+p-1}\left(\frac{1}{|J|}\sum_{j\in J}W_j^{(k)}\right)^{-1}. 
\end{equation}
Note that $\Sigma_{k+1}$ given above depends on $\nu$.

For $\Omega_{k+1}$ we obtain
\begin{equation}
\Omega_{t}=\frac{n+p+m-1}{p}\frac{1}{N}\sum_{i}^{N}(T_{i}-M_{t})W_{i}(T_{i}-M_{t})
\end{equation}
and for $n_{t}$, in the case where $\Sigma_{t}$ is a free parameter
that needs to be estimated, we have:
\begin{equation}
-p\log(m+n+p-1)+p\log(n_{t}+p-1)+\frac{1}{N}\sum_{i}^{N}\log|W_{i}|-\log\det\left|\frac{1}{N}\sum_{i}^{N}W_{i}\right|+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
\end{equation}
otherwise if $\Sigma_{t}=\Sigma$ is fixed and does not depend on
$n_{t}$:
\begin{equation}
\log\det\Sigma_{t}-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
\end{equation}
